---
title: "Project Title"
subtitle: "INFO 526 - Summer 2024 - Final Project"
author: 
  - name: "Team name"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project description"
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
---

## Abstract

Add project abstract here.

### Plot Setup

```{r setup}
# load packages
library(tidyverse)
library(glue)
library(here)
library(countdown)
library(ggthemes)
library(gt)
library(openintro)
library(ggrepel)
library(patchwork)

# set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))

# set width of code output
options(width = 65)

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7,        # 7" width
  fig.asp = 0.618,      # the golden ratio
  fig.retina = 3,       # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300             # higher dpi, sharper image
)

if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(tidyverse,
               glue,
               scales,
               ggthemes) 

devtools::install_github("tidyverse/dsbox")

```

## Introduction - dataset 1

Introduction: The introduction provides a clear explanation of the question and the dataset used to answer the question, including a description of all relevant variables in the dataset. (5 points) Meredith

## Justification of approach - data set1

The chosen analysis approach and visualizations are clearly explained and justified. (4 points) Stacey

## Code and Visualization - data set1

Code is correct, easy to read, properly formatted, and properly documented. (2.5 points) Brooke 
The visualizations are appropriate, easy to read, and properly labeled. (5 points) Brooke

## Code block 1

```{r Arizona crime plots}

# Load and remove meta data from the first 4 rows
az_crimes <- read_csv(here("data", "Arizona_Offense_Type_by_Agency_2023.csv"), skip = 4, show_col_types = FALSE)

# Select only city and crime columns of interest and rename
az_clean <- az_crimes %>%
  select(
    City = ...2,
    Crimes_Against_Persons = `Crimes Against Persons`,
    Crimes_Against_Property = `Crimes Against Property`,
    Crimes_Against_Society = `Crimes Against Society`
  ) %>%
  # Remove rows where City is NA or empty
  filter(!is.na(City) & City != "") %>%
  # Convert crime columns to numeric
  mutate(across(starts_with("Crimes_"), ~ as.numeric(str_replace_all(., ",", "")))) %>%
  # Remove rows where all crime counts are NA or zero
  filter(rowSums(across(starts_with("Crimes_"), ~ ifelse(is.na(.), 0, .))) > 0)

```

## Discussion 2

Discussion of results is clear and correct, and it has some depth without begin excessively long. (5 points) Meredith

## Introduction - dataset 2

Introduction: The introduction provides a clear explanation of the question and the dataset used to answer the question, including a description of all relevant variables in the dataset. (5 points) Meredith


## Justification of approach - data set 2

The chosen analysis approach and visualizations are clearly explained and justified. (4 points) Stacey

## Code and Visualization - data set 2

Code is correct, easy to read, properly formatted, and properly documented. (2.5 points) Brooke 
The visualizations are appropriate, easy to read, and properly labeled. (5 points) Brooke

## Discussion 2

Discussion of results is clear and correct, and it has some depth without begin excessively long. (5 points) Meredith 

## Sources

Referenced to use across function:
https://dplyr.tidyverse.org/reference/across.html
